HFI in RISC-V

### 2025-05-01 paper.tex
Encoding Instructions

@@

\section{Approach}
Our work brings HFI concepts to the RISC-V ecosystem by adapting the existing HFI design while fitting into RISC-V's existing privileged architecture. RISC-V presents both unique opportunities and challenges for HFI implementation. Its extensible ISA and clean-slate design allow for elegant integration of new instructions, while its simpler microarchitecture requires careful consideration of implementation details.

@@

@@

\section{Implementation \& Evaluation}
\subsection{Implementation - Intial Setup}
Since we were implenting our RISC-V HFI extension in QEMU, we cloned the current open-source QEMU repository and began work there. We began by taking time to understand how QEMU works, the purpose of the various repositories, and gaining a better understanding of RISC-V. We worked with the assumption that the system being run on always supports HFI but it is an easy addition to add any necessary hardware checks to validate that assumption.

@@

We tested the implementation using this single explicit data region. One obvious limitation is how we can scale this to work for multiple data regions, which can be a challenge due to the opcode limitation discussed in section 4. However, based on the feasibility of our implementation for the one data region and overall structure of the implementation, we believe proposed ideas in section 6 can allow easy extension to support more explicit data regions.

@@

\section{Future Work}

% TODO: Maybe move
\begin{table}[b]
  \centering
  \caption{Instruction encoding for \texttt{lw} in RISC-V}
  {\fontsize{6.5}{7.5}\selectfont
  \begin{tabular}{|c|c|c|c|c|}
  \hline
  \textbf{31--20} & \textbf{19--15} & \textbf{funct3} & \textbf{11--7} & \textbf{OPCODE} \\
  \hline
  \texttt{imm[11:0]} & \texttt{rs1} & \texttt{010} & \texttt{rd} & \texttt{0000011} \\
  \hline
  \end{tabular}
  }
  \label{tab:lw-encoding}
  \end{table}

  \begin{table}[b]
  \centering
  \caption{Instruction encoding for \texttt{sw} in RISC-V}
  {\fontsize{6.5}{7.5}\selectfont
  \begin{tabular}{|c|c|c|c|c|c|}
  \hline
  \textbf{31--25} & \textbf{24--20} & \textbf{19--15} & \textbf{funct3} & \textbf{11--7} & \textbf{OPCODE} \\
  \hline
  \texttt{imm[11:5]} & \texttt{rs2} & \texttt{rs1} & \texttt{010} & \texttt{imm[4:0]} & \texttt{0100011} \\
  \hline
  \end{tabular} 
  }
  \label{tab:sw-encoding}
  \end{table}

\subsection{Expand Number of Explicit Regions}
Previously mentioned, we currently implemented only a singular explict region with regards to the number of instructions that we would introduce.
In the simplest of implementations, we would just associate a separate instruction for each explicit region, but as we have discussed in section 4, we only have a finite number of opcodes.

@@

>>>>

\usepackage{float}

@@

\section{Approach}
\label{sec:approach}

Our work brings HFI concepts to the RISC-V ecosystem by adapting the existing HFI design while fitting into RISC-V's existing privileged architecture. 
RISC-V presents both unique opportunities and challenges for HFI implementation. Its extensible ISA and clean-slate design allow for elegant integration of new instructions, while its simpler microarchitecture requires careful consideration of implementation details.

@@

\subsection{Custom Instruction Encoding in RISC-V}
Before we consider custom instructions, we will first consider the key details of instruction encoding in RISC-V.
For starters, RISC-V is a fixed instruction length architecture with two variant sizes 16-bit (compressed) and 32-bit.
For our purposes, we will only need to consider the 32-bit instructions.

% TODO put in better spot
\begin{table}[H]
  \centering
  \caption{RISC-V Instruction Format Encodings}
  {\fontsize{6.5}{7.5}\selectfont
  \begin{tabular}{|c|c|c|c|c|c|c|}
  \hline
  \textbf{31--25} & \textbf{24--20} & \textbf{19--15} & \textbf{14--12} & \textbf{11--7} & \textbf{6--0} & \textbf{Type} \\
  \hline
  \texttt{funct7} & \texttt{rs2} & \texttt{rs1} & \texttt{funct3} & \texttt{rd} & \texttt{opcode} & R-type \\
  \hline
  \multicolumn{2}{|c|}{\texttt{imm[11:0]}} & \texttt{rs1} & \texttt{funct3} & \texttt{rd} & \texttt{opcode} & I-type \\
  \hline
  \texttt{imm[11:5]} & \texttt{rs2} & \texttt{rs1} & \texttt{funct3} & \texttt{imm[4:0]} & \texttt{opcode} & S-type \\
  \hline
  \end{tabular} 
  }
  \label{tab:encoding-summary}
\end{table}

Although Table~\ref{tab:encoding-summary} is not a complete representation of all the types of instructions in RISC-V, it will suffice for our implementation.
Register-type (R-type) denotes instructions that function entirely on registers such as \textit{add} and \textit{or}.
Immediate-type (I-type) are instructions that will require an immediate such as \textit{lw} (load word) and \textit{addi}.
Store-type (S-type) are specified for specifically stores to memory where the immediate is split into two parts; an instruction of this variant is \textit{sw} (store word).
For our purposes R-type will be utilized for updating and retrieving state of the sandbox such as the size of the implicit regions, base and bound of explicit regions, and their associated permissions.
I-type and S-type will be in regards to the loads and stores for the explicit regions with the h-prefixed instructions mentioned in Section~\ref{sec:impl-eval}.

In each of these instruction types, certain fields remain consistent, these are the first source register (rs1), the funct3 field, and the opcode. 
The funct3 and opcode fields are used together to distinguish between specific instructions. 
The opcode generally defines the broad category of the instruction, while funct3 provides finer-grained differentiation within that category.

\begin{table}[H]
  \centering
  \caption{Encoding of RISC-V Load Instructions}
  \small
  {
  \begin{tabular}{|c|c|c|c|c|c|}
    \hline
    \textbf{imm} & \textbf{rs1} & \textbf{funct3} & \textbf{rd} & \textbf{opcode} & \textbf{Instruction} \\
    \hline
    \texttt{imm[11:0]} & \texttt{rs1} & \texttt{000} & \texttt{rd} & \texttt{0000011} & \texttt{lb} \\
    \texttt{imm[11:0]} & \texttt{rs1} & \texttt{001} & \texttt{rd} & \texttt{0000011} & \texttt{lh} \\
    \texttt{imm[11:0]} & \texttt{rs1} & \texttt{010} & \texttt{rd} & \texttt{0000011} & \texttt{lw} \\
    \texttt{imm[11:0]} & \texttt{rs1} & \texttt{100} & \texttt{rd} & \texttt{0000011} & \texttt{lbu} \\
    \texttt{imm[11:0]} & \texttt{rs1} & \texttt{101} & \texttt{rd} & \texttt{0000011} & \texttt{lhu} \\
    \texttt{imm[11:0]} & \texttt{rs1} & \texttt{110} & \texttt{rd} & \texttt{0000011} & \texttt{lwu} \\
    \texttt{imm[11:0]} & \texttt{rs1} & \texttt{011} & \texttt{rd} & \texttt{0000011} & \texttt{ld} \\
    \hline
  \end{tabular}
  }
  \label{tab:load-instructions}
\end{table}

For example, in Table~\ref{tab:load-instructions}, see that their opcode is the same indicating their broad category of load and the funct3 value is varied for the specialized instruction.

Currently, there are 4 opcodes that are available for extending the instruction set; these are custom-0 (0001011), custom-1 (0101011), custom-2 (1011011), and custom-3 (1111011).
It is important to note that the opcodes custom-2 and custom-3 have been reserved for the open proposal for RV128, which will specify the RISC-V ISA for 128-bit machines. 
However, these remain available in RV32 and RV64. 
For simplicity, we will focus our efforts on extending RV64 to ensure no potential conflicts, but modifications should be considered to reduce the number of used opcodes and instructions to maximize compatibility.

In the composition of new instructions that interact with the underlying HFI state using the R-type instructions, we will use the custom-0 instruction.
Additionally, if we return to Table~\ref{tab:encoding-summary}, we can see that these R-type instructions also have a funct7 value which can be used to further increase the number of instructions with a fixed value for funct3 and opcode. 
Therefore, further consolidation can be performed for many of the R-type instructions. 
However, it should be mentioned that in creating instructions to expand functionality such as in Section~\ref{sec:impl-eval}, we did not use this as a focus, and determined this could be determined later.
As a result, our chosen funct3 and funct7 values can be consolidated within many of our implemented instructions.

Furthermore, it can be seen that loads and stores have little room for expansion with the absence of a funct7 value, and for the explicit region problems arise, which are discussed in Section~\ref{sec:future-work}.

\subsection{Extending QEMU for HFI}

\subsection{Region Enforcement Semantics}

@@

\section{Implementation \& Evaluation}
\label{sec:impl-eval}

\subsection{Implementation - Intial Setup}
Since we were implenting our RISC-V HFI extension in QEMU, we cloned the current open-source QEMU repository and began work there. We began by taking time to understand how QEMU works, the purpose of the various repositories, and gaining a better understanding of RISC-V. We worked with the assumption that the system being run on always supports HFI but it is an easy addition to add any necessary hardware checks to validate that assumption.

@@

We tested the implementation using this single explicit data region. One obvious limitation is how we can scale this to work for multiple data regions, which can be a challenge due to the opcode limitation discussed in Section~\ref{sec:approach}. However, based on the feasibility of our implementation for the one data region and overall structure of the implementation, we believe proposed ideas in Section~\ref{sec:future-work} can allow easy extension to support more explicit data regions.

@@

\section{Future Work}
\label{sec:future-work}

\subsection{Expand Number of Explicit Regions}
Previously mentioned, we currently implemented only a singular explict region with regards to the number of instructions that we would introduce.
In the simplest of implementations, we would just associate a separate instruction for each explicit region, but as we have discussed in Section~\ref{sec:approach}, we only have a finite number of available opcodes.

% TODO: Maybe move
\begin{table}[H]
  \centering
  \caption{Instruction encoding for \texttt{lw} in RISC-V}
  {\fontsize{6.5}{7.5}\selectfont
  \begin{tabular}{|c|c|c|c|c|}
  \hline
  \textbf{31--20} & \textbf{19--15} & \textbf{funct3} & \textbf{11--7} & \textbf{opcode} \\
  \hline
  \texttt{imm[11:0]} & \texttt{rs1} & \texttt{010} & \texttt{rd} & \texttt{0000011} \\
  \hline
  \end{tabular}
  }
  \label{tab:lw-encoding}
  \end{table}

  \begin{table}[H]
  \centering
  \caption{Instruction encoding for \texttt{sw} in RISC-V}
  {\fontsize{6.5}{7.5}\selectfont
  \begin{tabular}{|c|c|c|c|c|c|}
  \hline
  \textbf{31--25} & \textbf{24--20} & \textbf{19--15} & \textbf{funct3} & \textbf{11--7} & \textbf{opcode} \\
  \hline
  \texttt{imm[11:5]} & \texttt{rs2} & \texttt{rs1} & \texttt{010} & \texttt{imm[4:0]} & \texttt{0100011} \\
  \hline
  \end{tabular} 
  }
  \label{tab:sw-encoding}
  \end{table}

@@

\bibitem{RISCVSpec}
“V Ratified Specifications.” RISC, riscv.org/specifications/ratified/. Accessed 30 Apr. 2025.
